---
title: Converting long to wide data in R
author: Carlos Rodriguez
date: '2022-02-13'
slug: long-to-wide
categories: []
tags: []
subtitle: ''
summary: ''
authors: []
lastmod: '2022-02-13T22:18:29-07:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
draft: false
type: book
weight: 15
---

In the previous guide, we took a wide data set and converted it to long. While this conversion allowed us to apply some statistical and plotting functions on the data, there may be cases where doing the reverse, going from long to wide, may be useful.

```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(AMCP)
library(kableExtra)

# Load the data
data(chapter_12_table_1)
```

<!-- Create the long data frame -->
```{r, echo = FALSE, include=TRUE, warning=FALSE, message=FALSE}
# Create a new data frame with a subject id
long.data <- cbind(id = c(1:10), chapter_12_table_1)

# Convert the data from wide to long and convert id, Condition and Angle to factor
long.data <-  long.data %>% 
  gather(key = Condition.Angle, 
         value = Reaction_Time, 
         -id) %>%
  separate(col = Condition.Angle, 
           into = c("Condition", "Angle"), 
           sep = -1) %>%
  arrange(id, 
          Condition, 
          Angle)

kable(head(long.data), booktabs = TRUE, caption = "Long data format.")
```

## Converting Long to Wide
### 1. Unite the Condition and Angle columns
To convert our long format data to the original wide format data frame, we will need to chain a series of commands in the reverse order. Recall that our original wide data frame started out with columns that had the Condition and Angle together and that we had to separate those two variables. We will do the reverse here with the `unite()` function. We will merge the Condition and Angle column data into one column called Condition.Angle. 
```{r}
wide.data <- 
  long.data %>%
  unite(Condition.Angle, Condition, Angle, sep = "")
```
```{r, echo=FALSE}
kable(head(wide.data), booktabs = TRUE, caption = "Joining the condition and angle column cells.")
```

### 2. Pivot the data
Next, we will use the `pivot_wider()` function to arrange the values from the Reaction_Time column into separate columns names by the variables in the Condition.Angle columns.
```{r}
wide.data <- 
  long.data %>%
  unite(Condition.Angle, Condition, Angle, sep = "") %>%
  pivot_wider(names_from = Condition.Angle, values_from = Reaction_Time)
```

```{r, echo=FALSE}
kable(wide.data, booktabs = TRUE, caption = "Wide data format.")
```

### 3. Remove the id column
Finally, I then selected all columns except id to end up with our original data frame.
```{r}
wide.data <- 
  long.data %>%
  unite(Condition.Angle, Condition, Angle, sep = "") %>%
  pivot_wider(names_from = Condition.Angle, values_from = Reaction_Time) %>%
  select(-id)
```
```{r, echo=FALSE}
kable(wide.data, booktabs = TRUE, caption = "Final data frame matching the original chapter_12_table_1 data.")
```

### References
<div id="refs" class="references">


<div id="ref-R-rstatix">

Kassambara, Alboukadel. 2020a. *Rstatix: Pipe-Friendly Framework for Basic Statistical Tests*. <https://CRAN.R-project.org/package=rstatix>.

</div>

<div id="ref-R-AMCP">

Maxwell, Scott, Harold Delaney, and Ken Kelley. 2020. *AMCP: A Model Comparison Perspective*. <https://CRAN.R-project.org/package=AMCP>.

</div>


<div id="ref-R-tidyverse">

Wickham, Hadley. 2019. *Tidyverse: Easily Install and Load the ’Tidyverse’*. <https://CRAN.R-project.org/package=tidyverse>.

</div>

</div>
